---
title: "Rays Data Project Notes"
author: "Cameron Poppen"
date: "2024-11-24"
output: html_document
---

## Data
First, we're going to load in the necessary library for functions and take a look at a basic summary of the data.

```{r data}
library(tidyverse)
data <- read.csv("battedBallData.csv")
summary(data)
```

## Editing Data

Now, that we have saved the data, we're going to edit it to get the results we want.

```{r mutate}
data <- data %>%
  mutate(
    true_exit_speed = case_when(
      !is.na(speed_A) & !is.na(speed_B) ~ 0.85 * speed_A + 0.15 * speed_B,
      !is.na(speed_A) ~ speed_A,
      !is.na(speed_B) ~ speed_B,
    )
  )
```

If both of the measurement systems had data for the specific hit, I decided to use a weighted average using both.

I took 85% of system A, since it's the most reliable, and use 15% of B, just in case there is some error in system A.

But if either of them is NA, I just keep the data as given.

## Save Data

I saved the batters average speed based on the "batter id" given and the average of the weighted averages of their batted balls

```{r save}
batter_avg_speed <- data %>%
  group_by(batter) %>%
  summarize(
    avg_speed = mean(true_exit_speed, na.rm = TRUE)
  )
batter_avg_speed
```

I chose this simple model because it's very difficult to perfectly predict a players performance the following year, but the best data we have is based on the previous years, all things considered.

## Player Evaluation

The following code will help is find how well a player compares to the rest of the players

```{r players}
mean_speed <- mean(batter_avg_speed$avg_speed, na.rm = TRUE)
sd_speed <- sd(batter_avg_speed$avg_speed, na.rm = TRUE)

mean_plus_2sd <- mean_speed + 2 * sd_speed
mean_plus_1sd <- mean_speed + sd_speed
mean_minus_1sd <- mean_speed - sd_speed
mean_minus_2sd <- mean_speed - 2 * sd_speed
```

The following code tells whether a player is great, good, above average, below average, bad, or terrible, basing it on how they fair on the standard deviations.

``` {r sd}
batter_avg_speed <- batter_avg_speed %>%
  mutate(
    projected_performance = case_when(
      avg_speed > mean_plus_2sd ~ 'Great',
      avg_speed > mean_plus_1sd ~ 'Good',
      avg_speed > mean_speed ~ 'Above average',
      avg_speed > mean_minus_1sd ~ 'Below average',
      avg_speed > mean_minus_2sd ~ 'Bad',
      TRUE ~ 'Terrible'
    )
  )
```

The rest of the code in the ensures that the output is fairly readable and understandable.